\section{Exporting from Agda to Haskell}

\subsection{Goal}

As of Agda 2.3.2.2 there is a way to call Haskell code from Agda but no way to call Agda code from Haskell:
i.e. to compile Agda code into a library usable from Haskell.

The feature is desirable because one can create a formally verified library in Agda, create a simple API
around it and then use it in Haskell for real world applications.

\subsection{Current state of things in Agda}

\subsubsection{As of Agda 2.3.2.2}

There is a compiler from Agda to Haskell called MAlonzo\cite{MAlonzo}(which is
a rewrite of Alonzo\cite{Alonzo} compiler) - it translates Agda code
into a Haskell code so that observable behaviour of running an Agda interpreter and generated Haskell code is the same.

MAlonzo focuses on generating executables and therefore it has some undesirable characteristics:
\begin{itemize}
\item Names for functions and datatypes are of the form: a letter + a magic number
\item Type signatures for functions are not written
\item Generated datatypes omit all type parameters
\end{itemize}

On the other hand, apart from datatype parameters nothing is omitted:
even arguments that represent proofs are carried(but are replaced with \texttt{()} type)

Still, at this state using generated code from Haskell is not pleasant.

\subsubsection{Since Agda 2.3.4}

New \texttt{\{-\# COMPILED\_EXPORT \(AgdaName\) \(HaskellName\) \#-\}} pragma is added.
It will force the MAlonzo compiler to use \(HaskellName\) instead of a generated name for \(AgdaName\) function and will
also generate an explicit type signature for \(HaskellName\) according to the rules:

\begin{align*}
   T\llbracket Set\ A \rrbracket &= Unit\\
   T\llbracket x\ As \rrbracket &= x\ T\llbracket As \rrbracket\\
   T\llbracket (x : A) \rightarrow B \rrbracket &= \left\{
      \begin{array}{l l}
         \forall x.  \ T\llbracket A \rrbracket
            \rightarrow T\llbracket B \rrbracket &\quad x \in freevars(B)\\
         T\llbracket A \rrbracket
            \rightarrow T\llbracket B \rrbracket &\quad otherwise
      \end{array}\right.\\
   T\llbracket k\ As \rrbracket &= \left\{
      \begin{array}{l l}
         T\ T\llbracket As \rrbracket &\quad \texttt{COMPILED\_TYPE}\ k\ T\\
         Unit &\quad \texttt{COMPILED}\ k\ E
      \end{array}\right.
\end{align*}

Also MAlonzo gained the ability to generate a Haskell library - not only an executable.

While improving the experience it still lacks better datatype generation.

\subsection{The scope of this work}

This work was developed on Agda 2.3.2.2 and features of 2.3.4 were not available.

The \texttt{\{-\# EXPORT \(AgdaName\) \(HaskellName\) \#-\}} pragma is introduced which
will create a wrapper named \(HaskellName\) for MAlonzo generated code for \(AgdaName\) function or datatype.

The primary motivation is exposing functions with such types that using them in any legitimate way
from Haskell(i.e. no \texttt{unsafeCoerce} and such) will not break any internal invariant in Agda
code. It automatically means that only functions with types that can be expressed in Haskell can be
exposed - otherwise we lose information and invariants as a consequence.

Also for convenience Church polymorphism is converted to Curry polymorphism:
instead of(as can be seen in Agda 2.3.4)
\[
   TT\llbracket (A : Set) \rightarrow B \rrbracket = \forall a.\ () \rightarrow TT\llbracket B \rrbracket
\]
this holds
\[
   TT\llbracket (A : Set) \rightarrow B \rrbracket = \forall a.\ TT\llbracket B \rrbracket.
\]

Overview of features:
\begin{enumerate}
\item Functions with types expressed in Haskell can be exported and will
      undergo Church to Curry polymorphism transformation.
\item Datatypes can be exported as abstract datatypes when their type
      parameters are of types equivalent to kinds in Haskell(a combination
      of \(Set_0\) and arrows).
\item Constructors of this datatypes can be exported as Haskell functions
      (i.e. loosing the ability to pattern match on them) via feature 1.
\item Some Agda primitives: \texttt{INTEGER}, \texttt{FLOAT}, \texttt{CHAR},
      \texttt{STRING}, \texttt{IO} will be exported as \(Int,\ Float,\ Char,\ String,\ IO\)
      respectively.
\item \texttt{COMPILED\_TYPE}s will be exported accordingly.
\end{enumerate}

Section \ref{sec:wrappers} gives formal definition of wrappers generated,
section \ref{sec:proofs} gives proofs that exposed interface will not break the system.

\section{Wrapper generation}\label{sec:wrappers}

\subsection{Kinds}
\[ KT\llbracket Kind \rrbracket = HaskellKind \]

\begin{align*}
   KT\llbracket Set_0 \rrbracket &= *\\
   KT\llbracket Kind_1 \rightarrow Kind_2 \rrbracket &=
      KT\llbracket Kind_1 \rrbracket \rightarrow KT\llbracket Kind_2 \rrbracket
\end{align*}

\subsection{Type declarations}

\(DTMA\) gives MAlonzo generated type name.

\(DT,\ DTD\) are defined when \texttt{\{-\# EXPORT} \(AgdaTypeName\) \(HaskellTypeName\) \texttt{\#-\}} is specified.

\begin{align*}
   &DTMA\llbracket AgdaTypeName \rrbracket = HaskellTypeName\\
   &DT\llbracket AgdaTypeName \rrbracket = HaskellTypeName\\
   &DTD\llbracket AgdaTypeName \rrbracket \doteq HaskellTypeDeclaration\\
\end{align*}

Considering declaration:
\begin{align*}
   &\textbf{data } AgdaDataType\ (A_1 : Kind_1) \cdots (A_m : Kind_m) :
      Kind_{m+1} \rightarrow \cdots \rightarrow Kind_n \rightarrow Set \textbf{ where } \ldots
\end{align*}

\begin{align*}
   &DTD\llbracket AgdaDataType \rrbracket \doteq\\
   &\quad\textbf{newtype } DT\llbracket AgdaDataType \rrbracket\ (a_0 :: KT\llbracket Kind_1 \rrbracket)
      \cdots (a_n :: KT\llbracket Kind_n \rrbracket)\\
   &\quad\quad = DT\llbracket AgdaDataType \rrbracket\ (\forall b_1 \cdots b_k.\ DTMA\llbracket AgdaDataType \rrbracket\ b_1\ \cdots\ b_k)
\end{align*}

\(k\) is an arity of type constructor generated by MAlonzo.

It also works for \textbf{record}s.

\subsection{Types}

\subsubsection{Primitive transformation}

Only primitives that are used for postulates are allowed.  MAlonzo defines:
\[
   PTMA\llbracket PrimitiveType \rrbracket = HaskellType
\]
\begin{align*}
   PTMA\llbracket \texttt{INTEGER} \rrbracket &= Int\\
   PTMA\llbracket \texttt{FLOAT} \rrbracket &= Float\\
   PTMA\llbracket \texttt{CHAR} \rrbracket &= Char\\
   PTMA\llbracket \texttt{STRING} \rrbracket &= String\\
   PTMA\llbracket \texttt{IO} \rrbracket &= IO
\end{align*}

\subsubsection{Type transformation}

\begin{align*}
   &TT\llbracket AgdaType \rrbracket(Context) = HaskellType\\
   &Context = \{ AgdaTypeVarName \mapsto HaskellTypeVarName \}
\end{align*}

\begin{align*}
   TT\llbracket T\ args\ldots \rrbracket(\Gamma) &= \left\{
      \begin{array}{l l}
         a\ TT\llbracket args\ldots \rrbracket(\Gamma) &\quad (T \mapsto a) \in \Gamma \\
         CT\ TT\llbracket args\ldots \rrbracket(\Gamma) &\quad
            \text{\texttt{\{-\# COMPILED\_TYPE \(T\) \(CT\) \#-\}} is specified}\\
         PTMA\llbracket P \rrbracket\ TT\llbracket args\ldots \rrbracket(\Gamma) &\quad
            \text{\(PTMA\llbracket P \rrbracket\) is defined and
            \texttt{\{-\# BUILTIN \(P\) \(T\) \#-\}} is specified}\\
         DT\llbracket T \rrbracket\ TT\llbracket args\ldots \rrbracket(\Gamma) &\quad
            \text{\(DT\llbracket T \rrbracket\) is defined}\\
      \end{array}\right.\\
   TT\llbracket (A : Kind) \rightarrow T \rrbracket(\Gamma) &= \forall (a :: KT\llbracket Kind \rrbracket).
      \ TT\llbracket T \rrbracket(\Gamma \cup \{A \mapsto a\})\\
   TT\llbracket (x : T_1) \rightarrow T_2 \rrbracket(\Gamma) &=
      TT\llbracket T_1 \rrbracket(\Gamma) \rightarrow TT\llbracket T_2 \rrbracket(\Gamma)
      \quad x \not\in freevars(T_2) \\
   TT\llbracket (x : T_1,\ T_2) \rrbracket(\Gamma) &=
      (TT\llbracket T_1 \rrbracket(\Gamma),\ TT\llbracket T_2 \rrbracket(\Gamma))
      \quad x \not\in freevars(T_2) \\
\end{align*}

\subsection{Terms}

\(Wrap^0\llbracket AgdaType \rrbracket(term)\) is defined only when \(TT\llbracket AgdaType \rrbracket(\varnothing)\) is defined.

\begin{align*}
   &Wrap^{2k}\llbracket AgdaType \rrbracket(MAlonzoTerm) = MyTerm\\
   &Wrap^{2k + 1}\llbracket AgdaType \rrbracket(MyTerm) = MAlonzoTerm
\end{align*}

\begin{align*}
   Wrap^k\llbracket A\ args\ldots \rrbracket(term) &= \texttt{unsafeCoerce } term\\
   Wrap^{2k}\llbracket (A : Kind) \rightarrow T \rrbracket(term) &= Wrap^{2k}\llbracket T \rrbracket(term\ ())\\
   Wrap^{2k+1}\llbracket (A : Kind) \rightarrow T \rrbracket(term) &= Wrap^{2k+1}\llbracket T \rrbracket(\lambda \_.\ term)\\
   Wrap^k\llbracket (x : T_1) \rightarrow T_2 \rrbracket(term) &=
      \lambda x.\ Wrap^k\llbracket T_2 \rrbracket(term\ Wrap^{k+1}\llbracket T_1 \rrbracket(x))\\
   Wrap^k\llbracket (x : T_1,\ T_2) \rrbracket((term_1,\ term_2)) &=
      (Wrap^k\llbracket T_1 \rrbracket(term_1),\ Wrap^k\llbracket T_2 \rrbracket(term_2))
\end{align*}

\subsection{Value declarations}

\(VTMA\) gives MAlonzo generated value name

\(VT,\ VTD\) are defined when \texttt{\{-\# EXPORT} \(AgdaName\) \(HaskellName\) \texttt{\#-\}} is specified.

\begin{align*}
   &VTMA\llbracket AgdaName \rrbracket = HaskellName\\
   &VT\llbracket AgdaName \rrbracket = HaskellName\\
   &VTD\llbracket AgdaName \rrbracket \doteq HaskellDeclaration\\
\end{align*}

Considering declaration:
\begin{align*}
   &AgdaName : AgdaType\\
   &AgdaName = \ldots
\end{align*}

\begin{align*}
   &VTD\llbracket AgdaName \rrbracket \doteq\\
   &\quad\quad VT\llbracket AgdaName \rrbracket :: TT\llbracket AgdaType \rrbracket(\varnothing)\\
   &\quad\quad VT\llbracket AgdaName \rrbracket = Wrap^0\llbracket AgdaType \rrbracket(VTMA\llbracket AgdaName \rrbracket)\\
\end{align*}

It works in the same way for constructors(it exports them as Haskell functions - not Haskell constructors).
It also works seamlessly with parametrized modules and, consequently, with record functions.

\section{Proofs}\label{sec:proofs}

\subsection{Old Intro}

I need to prove that Haskell types and terms that I expose wouldn't break the system.  It means two things:
\begin{enumerate}
\item Types preserve the same set of invariants
\item Terms have the same interface: any combination of \texttt{APPLY} that can be used(ignoring types) to original
      term must be usable with generated one; and primitives(numbers, strings, ... and their ops) are the same.
\end{enumerate}

\subsection{Preserving type invariants}

Three cases:
\begin{enumerate}
\item \textbf{newtype} wrappers.

A datatype can be viewed as a logical statement and its constructors --- as
proofs of this statement. \textbf{newtype} wrapping gives us a statement but
hides proofs. Acquiring an instance of this datatype in Haskell can only be done
when some Agda function returns it. Therefore it was constructed with all the invariants
checked by Agda. We can only use this datatype with functions exported from Agda that
used the original version of it. Therefore all the internal invariants are safely passed
from Agda to Agda invisibly through Haskell.

\item Transformation from Church polymorphism to Curry polymorphism.

\begin{align*}
&(A : Kind) \rightarrow Type \\
&(\forall a :: KT\llbracket Kind \rrbracket).\ HaskellType
\end{align*}

They both mean the same thing but the first one always requires a proof that
\(Kind\) is inhabited:

\begin{itemize}
\item If \(A \not\in freevars(Type)\) and, by construction(\(TT\)), \(a \not\in freevars(HaskellType)\).
\item \(A\) (and consequently \(a\)) is a phantom type(i.e. only used as a type parameter).
\end{itemize}

In both cases Haskell will completely ignore the inhabitance of \(KT\llbracket Kind \rrbracket\).
Agda however will require you to provide an evidence that \(Kind\) can be constructed. Now,
\(Kind\) is defined as a combination of \(Set_0\) and arrows. Therefore some \(Kind\ A\) can be viewed as
follows: \(Arg_1 \rightarrow \ldots \rightarrow Arg_n \rightarrow Set_0\) for \(n \ge 0\).
Let's define a simple \(Unit\) type:
\begin{align*}
   &\mathbf{data}\ Unit : Set\ \mathbf{where}\\
   &\quad unit : Unit
\end{align*}

We can now construct an \(A\): \(A = \lambda arg_1 \ldots arg_n.\ Unit\). Therefore, each \(Kind\)
is inhabited and we can safely omit this proof in our transformation.

\item In every other case type is exactly the same.

So invariants are clearly the same.
\end{enumerate}

\subsection{Preserving term interface}

\(Wrap\) clearly deals with the issue of passing and skipping type parameters with
MAlonzo-generated code.

A thing to watch for is \texttt{unsafeCoerce}.  There are three cases for a coerced type:
\begin{enumerate}
\item a \textbf{newtype} wrapper around an MAlonzo-generated datatype.

   Safe because \textbf{newtype} is required to have the same
   internal structure as its wrapped type.

\item a primitive as defined by \(PTMA\).

   Safe because type of MAlonzo-generated code is the same as ours by construction.

\item \(a\ args\ldots\), where \(a\) is a type variable.

   Safe because all terms with type \(a\ args\ldots\) will have the same internal
   structure. That's because from Haskell side compiler will guarantee that and
   from Agda side terms will have a corresponding type \(A\ args\ldots\) (via \(\Gamma\) in \(TT\))
   so the compiler will guarantee it too.
\end{enumerate}

\begin{thebibliography}{9}
\bibitem{MAlonzo}
   Makoto Takeyama,
   \emph{a new compiler MAlonzo}.
   \url{http://thread.gmane.org/gmane.comp.lang.agda/62},
   2008.

\bibitem{Alonzo}
   Marcin Benke,
   \emph{Alonzo --- a compiler for Agda}.
   TYPES2007,
   2007.
\end{thebibliography}
